---
title: "Fit Income Distribution from Gini"
author: "Ben Mazzotta"
date: "Wednesday, October 15, 2014"
output:   pdf_document
header-includes:
      - \usepackage{times}
      - \usepackage{euler}
---


## Overview

Personal income data can be fit most accurately to a lognormal distribution of income, when calibrating to the Gini coefficient on income (Eric Kemp-Benedict 2001). The evidence for this is quantile regression of the Lorenz curve that generates the Gini coefficient, when compared to similar analysis for the gamma, Pareto and beta distributions. The lognormal distribution has a distinct advantage over polynomial distributions in that it is characterized by just two parameters, $\mu$ and $\sigma$.



```{r prelims, results='hide'}
require(data.table); require(plyr); require(ggplot2)
setwd("../data")
load("current.Rdata")
```

## Model

Kemp-Benedict (2001:7) shows that the Gini index depends only on the parameter $\sigma$ of the lognormal distribution.

$$ g = 2 \Phi \left(  \frac{\sigma}{\sqrt 2} \right) -1 $$ 

... where $\Phi$  is the cumulative standard normal distribution. If we knew the value of $\sigma$, we could write the following code.

`g <- 2 * dnorm(sigma/sqrt(2)) - 1`

And hence to calculate $\sigma$ from a known Gini coefficient, we would invert as follows.

$$ \sigma = \sqrt{2} \Phi^{-1} \left( \frac{g+1}{2} \right) $$ 

The corresponding R code is:

`sigma = sqrt(2) * qnorm( (g+1)/2 )`



## Parameters

The scale and location parameters of the lognormal distribution are $\mu$ and $\sigma$. From Kemp above we can calculate $\sigma$.

We also know that average income is a function of  $\mu$ and $\sigma$. So inverting, we can derive:

$$E(I) = \exp{\mu + \frac{\sigma^2}{2}}$$

$$ \mu = log(E(I)) - \frac{\sigma^2}{2}$$

And hence the procedure should be to calculate:

1. Sigma from Gini coefficient
2. Mu from GNI/capita and sigma
3. Quantiles from known distribution I = f(mu, sigma; X).

```{r fitdistr}
inc <- inc2010[!is.na(gini),]
setkey(inc, iso2c, country, year)

##      Step 1: calculate sigma from Gini coefficient. 
##          (BDM: is this possible w/o scale parameter?)
inc[,sigma:= sqrt(2)*qnorm((gini/100+1)/2)]
summary(inc)
##      Step 2: calculate mu from sigma and mean
inc[,mu:=log(inccap)-(sigma^2/2)]
summary(inc)
##      Step 3: Read quantiles from mu and sigma
inc[,p20:=qlnorm(.20, mu, sigma)]
inc[,psav:=qlnorm(1-savers/100, mu, sigma)]

options(digits=4)
summary(inc)
summary(1-inc$savers/100)
```


So the object `inc` in memory has accurate parameters for the lognormal distribution of per capita income in memory.

It also has the 20th percentile of per capita income, and the nth percentile of per capita income, where $n$ corresponds to the share of the population that do not save at a formal financial institution.

Are these estimates plausible?

```{r plotincome, fig.width=8, fig.height=4.5}
## Plot p20 vs income per capita.
##      Has to be all below 45 degree line
qplot(inccap, p20, data=inc, geom="point", log="xy")+geom_abline(intercept=0,slope=1)+
  labs(list(title="Income levels: bottom quintile vs. average", x="Income per capita, USD at PPP", y="Income cutoff at the bottom quintile"))

## Plot psavers vs income per capita.
##      Rich countries below the line, poor countries above
qplot(inccap, psav, data=inc, geom="point", log="xy")+geom_abline(intercept=0,slope=1)+
  labs(list(title="Income levels: financially excluded proportion vs. average", x="Income per capita, USD at PPP", y="Income cutoff among financially excluded"))


```


## Simulate distributions

Now let's take the distribution parameters mu and sigma for each country. Using a large-enough population, simulate the share of income accruing to the bottom quintile of the distribution.

```{r simulateincome}
# inc.siminput <- inc[,.SD, .SDcols=c("mu","sigma")]

## Make a list of simulated income data
##      Each of length 10^6, using the observed values of mu and sigma
sim <- mapply(rlnorm, rep(1e6, nrow(inc)), inc$mu, inc$sigma)
##      It's very large.
dim(sim)
object.size(sim)/1e6
##      It took several seconds to compute the summary statistics.
# summary(sim)
sim <- data.table(sim)

sim.sums <- colSums(sim)

str(inc$p20)
head(inc$p20)

# ##    How to use data.table to get fast sums of subsets?
# sim[V1<235,sum(V1)]
# ## You can run this code to get the sum of bottom quintile values in a single column
# myvar <- "V1"
# sim[get(myvar)<234.4945, sum(.SD), .SDcols=myvar]
# # Can I rewrite so that only number 1 is an input?
# sim[get(paste("V",1,sep=""))<inc$p20[1], sum(.SD), .SDcols=paste("V",1,sep="")]

## Write a function to do this.
calc.p20sum <- function(x) sim[get(paste("V",x,sep=""))<inc$p20[x], sum(.SD), .SDcols=paste("V",x,sep="")]

# Does it still work? Yes. :)
calc.p20sum(1)

sim.sump20 <- sapply(c(1:length(inc$p20)), calc.p20sum)
summary(sim.sump20)

sim.p20share <- sim.sump20/sim.sums
sim.p20share *100

## And to do the same thing for the proportion saving:
  
## This function returns the gross income beneath the nth percentile of income,
##      Where n is the percentage of those that do not save.
calc.finexsum <- function(x) sim[get(paste("V",x,sep=""))<inc$psav[x], sum(.SD), .SDcols=paste("V",x,sep="")]

# Does it still work? Yes. :)
calc.finexsum(1)
# calc.finexsum(17)

# inc$psav

## Running the same command as above threw an error
##      Not all the psav observations are defined.
##      Need to substitute something when undefined.....
sim.sumfinex <- sapply(which(!is.na(inc$psav)), calc.finexsum)
summary(sim.sumfinex)

sim.finexshare <- rep(NA, nrow(inc))
sim.finexshare[which(!is.na(inc$psav))] <- sim.sumfinex / sim.sums[which(!is.na(inc$psav))]

# options(digits=3)
sim.finexshare*100

gini.dt <- inc
gini.dt[,bottom20share <- sim.p20share]
gini.dt[,finexshare <- sim.finexshare]

save(gini.dt, file="../data/income share of financially excluded.Rda")
rm(sim)
save.image("../data/working.Rdata")


```




## The lognormal distribution

Key features of the lognormal distribution

$$ X \sim lornorm(\mu, \sigma) $$ 

$$ E(X) = exp(\mu + \frac{\sigma^2}{2}) = e ^{\mu + \frac{\sigma^2}{2}} $$

$$ X_{median} = e^{\mu} $$

And hence we can calculate from average income per capita:

$$ \mu = ln(inccap) - \frac{\sigma^2}{2}$$

## Data

This is a dataset that uses sloppy averaging from World Development Indicators for proof of concept (world Bank 2014). 

Var | Data | Notes
--- | --- | ---
iso2c | 2-letter abbreviation | Provided by the Bank, supposedly ISO format.
country | Country name | 
year | Year | Grossly inaccurate!
income | National income | Current US$
incap | Per capita income | Current US$
gini | Gini index | A pure number on the (0,1 interval)
saving | National saving  | Current US$
savingrate | Saving rate | Percent of GNI
savers | Currently save with a bank | Percent of population

This is a summary of the most recent data available.

```{r, echo=FALSE}
str(inc2010)
summary(inc2010)
```

$\square$

## Bibliography

Eric Kemp-Benedict. 2001. Income Distribution and Poverty Data: Methods for Using Available Data in Global Analysis. Polestar Technical Note No. 4.

World Bank. 2014. World Development Indicators.


---
\footnotesize Ben Mazzotta is a postdoc investigating the [Cost of Cash](http://fletcher.tufts.edu/costofcash) at the [Institute for Business in the Global Context](http://fletcher.tufts.edu/ibgc) (IBGC). Contact: [benjamin.mazzotta@tufts.edu](mailto:benjamin.mazzotta@tufts.edu) \quad +1.617.627.4561 \quad 160 Packard Ave, Medford MA 02155. \par
