---
title: "Simulating quantile distributions"
author: "Ben Mazzotta"
date: "Wednesday, November 19, 2014"
output: pdf_document
header-includes: 
  - \usepackage{charter}
  - \usepackage{euler}
---

This is a proof of concept document for simulating quantile shares of a random variable.

**Problem.** What share of X do the bottom two quintiles possess?

```{r incomeshare}
set.seed(1213)
t.mu <- 8
t.sigma <- 2
qlnorm(.6, t.mu, t.sigma)
t.draw <- rlnorm(1e6, t.mu, t.sigma)
str(t.draw)
t.under40 <- subset(t.draw, t.draw<qlnorm(.40, t.mu, t.sigma))
sum(t.draw)
sum(t.under40)
sum(t.under40) / sum(t.draw)

```


Replicate for various sample sizes.

```{r listofincomeshares}
# ## Declare a list of quantiles to cutoff
# t.quant = c(.2, .4, .6, .8)

## Declare a list of sample sizes
t.sample = c(1e4, 1e5, 1e6)

## Simulate the distribution with three different sample sizes; save to list
t.draw2 <- lapply(t.sample, function(x) rlnorm(x, t.mu, t.sigma))
## Make a data frame of summary statistics
##      Column 1: sample size
t.df <- data.frame(t.sample)
##      Column 2: sum of X
t.df$sum <- sapply(t.draw2, sum)
##      Column 3: sum of X from bottom two deciles
t.df$sum40 <- sapply(t.draw2, function(x) sum(subset(x, x < qlnorm(.4, t.mu, t.sigma))))
##      Column 4: share of X belonging to bottom two deciles.
t.df$share40 <- t.df$sum40 / t.df$sum

cat("The bottom two quintiles of X are", t.df[3,4]*100, "% of the total.")

```



---
\footnotesize Ben Mazzotta works on the [Cost of Cash](http://fletcher.tufts.edu/costofcash) project at the [Institute for Business in the Global Context](http://fletcher.tufts.edu/ibgc). Contact: [benjamin.mazzotta@tufts.edu](mailto:benjamin.mazzotta@tufts.edu) \quad +1.617.627.4561 \quad 160 Packard Ave, Medford MA 02155. \par

